A general Nullstellensatz for generalized spaces

### 2019-03-11 paper-qcoh.tex
More careful rendition of contexts

To verify that~(3) implies~(4), we can mimick the usual proof of this fact in
the internal language of~$\E$: One can construct a variant of the syntactic
site of~$\TT$ which is built using only coherent sequents and finitary
derivability~\cite[Section~1.4]{caramello:tst}. The sheaf topos over this site
is another model for the classifying topos of~$\TT$. If~$\sigma$
is~$\TT$-derivable by a proof tree of arbitrary shape, then~$\sigma$ holds in
this model of the classifying topos. By the analogue of
Proposition~\ref{prop:classifying-topos} for this model, we obtain
that~$\sigma$ is~$\TT$-derivable by a proof tree of finite shape.

@@

\begin{lemma}\label{lemma:locally-constant}
Let~$\TT$ be a geometric theory.
Let~$\varphi$ be a section of the sheaf of geometric$^\star$ formulas over the signature
of~$\ul{\TT}/U_\TT$ over a stage~$A$. Then there is a covering~$(A_i \to A)_i$
of~$A$ such that for each index~$i$, there is a formula~$\varphi_i$ over the signature
of~$\TT/U_\TT(A_i)$ such that~$A_i \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
$(\varphi \dashv\vdash \varphi_i)$}$.
\end{lemma}

\begin{proof}By passing to a covering, we may suppose that~$\varphi$ is given
by an (external) geometric formula over the signature
of~$\ul{\TT}(A)/U_\TT(A)$. Any function symbol and relation symbol
of~$\ul{\TT}(A)$ occuring in~$\varphi$ is locally given by a symbol in~$\TT$.
Hence the claim would be trivial if~$\varphi$ were a coherent formula, for in
this case we would just have to pass to further coverings, one for each
occuring symbol, a finite number of times in total.

However, in general, we cannot conclude as easily. Write~$A = \{ \vec x\_
\alpha \}$. Let~$R$ be a relation symbol of~$\ul{\TT}(A)$ occuring
in~$\varphi$. Then there is a covering~$(\{ \vec y_j\_ \alpha_j \}
\xrightarrow{[\theta_j]} \{ \vec x\_ \alpha \})_j$ such that, restricted to~$\{
\vec y_j\_ \alpha_j \}$, $R$ is given by a symbol~$R_j$ of~$\TT$. To construct
the desired formula~$\varphi'$, we replace any such occurrence~$R(\ldots)$
in~$\varphi$ by
\[ \bigvee_j \bigl((\exists \vec y_j\_ \theta_j) \wedge R_j(\ldots)\bigr). \]
In a similar vein we treat any occurrence of function symbols.

The verification of~$A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
$(\varphi \dashv\vdash \varphi')$}$ rests on the observation
\[ A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\bigl((\exists \vec y_j\_
\theta_j) \seq{\vec x} \bigvee_j \{ \top \,|\, \text{$(\exists \vec y_j\_ \theta_j)$
holds for~$U_\TT$} \}\bigr)$} \]
which in turn follows from Lemma~\ref{lemma:truth-to-provability} exploiting
that~$(\exists \vec y_j\_ \theta_j) \wedge (\exists \vec y_k\_ \theta_k)
\seq{\vec x} \bigvee \{ \top \,|\, j = k \}$.
\end{proof}

@@

For the direction~$(1) \Rightarrow (2)$ we have to verify that, given any
stage~$A$ and any section~$\sigma$ of the sheaf of
geometric$^\star$ sequents over~$A$, if~$A \models
\speak{$\sigma$ holds for~$U_\TT$}$ then~$A \models
\speak{$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$}$. By
Lemma~\ref{lemma:locally-constant} we may suppose that~$\sigma$ is an
(external) geometric sequent over the signature of~$\TT/U_\TT(A)$.

@@

% XXX: check contexts.

@@

For the ``only if'' direction, we use that a statement holds in~$\Set[\TT']$ if
and only if its~$\nabla$-translation holds in~$\Set[\TT]$, where~$\nabla$ is
the modal operator associated to the topology cutting
out~$\Set[\TT']$~\cite[Theorem~6.31]{blechschmidt:phd}. Exploiting some of the
simplification rules of the~$\nabla$-translation~\cite[Section~6.6]{blechschmidt:phd},
it hence suffices to verify, internally to~$\Set[\TT]$, that:
\begin{multline*}
  \text{For any geometric$^\star$ sequent~$\sigma = (\varphi \seq{\vec x} \psi)$ where~$\psi$ is a Horn formula,} \\
    \text{if~$\forall x_1,\ldots,x_n\?U_\TT\_ (\varphi \Rightarrow \nabla
    \psi)$, then~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}
\end{multline*}
Since~$\nabla$ commutes with finite conjunctions and since the sheaf assumption
implies that $\nabla(s = t)$ is equivalent
to~$s = t$ and that, for relation symbols~$R$, the
statement~$\nabla(R(s_1,\ldots,s_m))$ is equivalent to~$R(s_1,\ldots,s_m)$,
the statement~$\nabla\psi$ is equivalent to~$\psi$. Hence the claim follows
from Theorem~\ref{thm:nullstellensatz}.
\end{proof}

@@

For the converse direction, we note that, given a geometric sequent~$\sigma$
over the signature of~$\TT$, the statement~$\speak{$\ul{\TT}/U_\TT$
proves$^\star$~$\sigma$}$ can be expressed as a geometric formula over the
signature of~$\TT$. Applying this observation successively to subformulas
of the given formula~$\alpha$, Theorem~\ref{thm:nullstellensatz} on the
semantic side and the axiom scheme~\eqref{eq:nullstellensatz} imply that we may
assume that~$\alpha$ is in fact a geometric formula. Hence we are reduced to
the basic fact that, for geometric formulas~$\varphi$, $\Set[\TT] \models
\varphi$ implies that~$\TT$ proves~$\varphi$.
\end{proof}

>>>>

To verify that~(3) implies~(4), we can mimic the usual proof of this fact in
the internal language of~$\E$: There is a variant of the syntactic
site of~$\TT$ which is built using only coherent sequents and finitary
derivability~\cite[Section~1.4]{caramello:tst}. The sheaf topos over this site
is another model for the classifying topos of~$\TT$, and still validates, like
any Grothendieck topos, full infinitary logic. Hence, if~$\sigma$
is~$\TT$-derivable by a proof tree of arbitrary shape, then~$\sigma$ holds in
this model of the classifying topos. By the analogue of
Proposition~\ref{prop:classifying-topos} for this model, we obtain
that~$\sigma$ is~$\TT$-derivable by a proof tree of finite shape.

@@

\begin{lemma}\label{lemma:locally-constant}
Let~$\TT$ be a geometric theory.
Let~$\varphi$ be a section of the sheaf of geometric$^\star$ formulas over the signature
of~$\ul{\TT}/U_\TT$ over a stage~$A \in \C_\TT$. Then there is a covering~$(A_i \to A)_i$
of~$A$ such that for each index~$i$, there is a formula~$\varphi_i$ over the signature
of~$\TT/U_\TT(A_i)$ such that~$A_i \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
$(\varphi \dashv\vdash \varphi_i)$}$.
\end{lemma}

\begin{proof}By passing to a covering, we may suppose that~$\varphi$ is given
by an (external) geometric formula over the signature
of~$\ul{\TT}(A)/U_\TT(A)$.

Any function symbol and relation symbol
of~$\ul{\TT}(A)$ occurring in~$\varphi$ is locally given by a symbol of~$\TT$.
Hence the claim would be trivial if~$\varphi$ were a coherent formula, for in
this case we would just have to pass to further coverings, one for each
occuring symbol, a finite number of times in total.

However, in general, we cannot conclude as easily. Write~$A = \{ \vec x\_
\alpha \}$. Let~$R$ be a relation symbol of~$\ul{\TT}(A)$ occuring
in~$\varphi$. By the explicit description of constant sheaves as sheaves of
locally constant maps, there is a covering~$(\{ \vec y_j\_ \alpha_j \}
\xrightarrow{[\theta_j]} \{ \vec x\_ \alpha \})_j$ such that, restricted to~$\{
\vec y_j\_ \alpha_j \}$, $R$ is given by a relation symbol~$R_j$ of~$\TT$. To construct
the desired formula~$\varphi'$, we replace any such occurrence~$R(\ldots)$
in~$\varphi$ by
\[ \bigvee_j \bigl((\exists \vec y_j\_ \theta_j) \wedge R_j(\ldots)\bigr). \]
In a similar vein we treat any occurrence of function symbols.

The verification of~$A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$
$(\varphi \dashv\vdash \varphi')$}$ rests on the observation
\[ A \models \speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\bigl((\exists \vec y_k\_
\theta_k) \seq{[]} \bigvee_k \{ \top \,|\, \text{$(\exists \vec y_k\_ \theta_k)$
holds for~$U_\TT$} \}\bigr)$} \]
which in turn can be checked on the covering~$(\{ \vec y_j\_ \alpha_j \}
\xrightarrow{[\theta_j]} \{ \vec x\_ \alpha \})_j$, applying
Lemma~\ref{lemma:truth-to-provability} and using that~$\TT$ (and
hence~$\ul{\TT}$) proves~$(\exists \vec y_j\_ \theta_j) \wedge (\exists \vec
y_k\_ \theta_k) \seq{\vec x} \bigvee \{ \top \,|\, j = k \}$.
\end{proof}

@@

For the direction~$(1) \Rightarrow (2)$ we have to verify that, given any
stage~$A \in \C_\TT$ and any section~$\sigma$ of the sheaf of
geometric$^\star$ sequents over~$A$, if~$A \models
\speak{$\sigma$ holds for~$U_\TT$}$ then~$A \models
\speak{$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$}$. By
Lemma~\ref{lemma:locally-constant} we may suppose that~$\sigma$ is an
(external) geometric sequent over the signature of~$\TT/U_\TT(A)$.

@@

@@

For the ``only if'' direction, we use that a statement holds in~$\Set[\TT']$ if
and only if its~$\nabla$-translation holds in~$\Set[\TT]$, where~$\nabla$ is
the modal operator associated to the topology cutting
out~$\Set[\TT']$~\cite[Theorem~6.31]{blechschmidt:phd}. Exploiting some of the
simplification rules of the~$\nabla$-translation~\cite[Section~6.6]{blechschmidt:phd},
it hence suffices to verify, internally to~$\Set[\TT]$, that:
\begin{multline*}
  \text{\emph{For any geometric$^\star$ sequent~$\sigma = (\varphi \seq{\vec x} \psi)$ where~$\psi$ is a Horn formula,}} \\
    \text{\emph{if~$\forall x_1,\ldots,x_n\?U_\TT\_ (\varphi \Rightarrow \nabla
    \psi)$, then~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}}
\end{multline*}
Since~$\nabla$ commutes with finite conjunctions and since the sheaf assumption
implies that $\nabla(s = t)$ is equivalent
to~$s = t$ and that, for relation symbols~$R$, the
statement~$\nabla(R(s_1,\ldots,s_m))$ is equivalent to~$R(s_1,\ldots,s_m)$,
the statement~$\nabla\psi$ is equivalent to~$\psi$. Hence the claim follows
from Theorem~\ref{thm:nullstellensatz}.
\end{proof}

@@

For the converse direction, we note that, given a geometric sequent~$\sigma$
over the signature of~$\TT$, the statement~$\speak{$\ul{\TT}/U_\TT$
proves$^\star$~$\sigma$}$ can be expressed as a geometric formula over the
signature of~$\TT$. Applying this observation successively to subformulas
of the given formula~$\alpha$, Theorem~\ref{thm:nullstellensatz} on the
semantic side and the axiom scheme~\eqref{eq:nullstellensatz} on the syntactic side imply that we may
assume that~$\alpha$ is in fact a geometric formula. Hence we are reduced to
the basic fact that, for geometric formulas~$\varphi$, $\Set[\TT] \models
\varphi$ implies that~$\TT$ proves~$\varphi$.
\end{proof}
